S-Aligner: Ultrascalable read mapping on Sunway Taihu Light

### 2017-04-07 text/design.tex
minor edits throughout

This section describes the implementation of both the inter-CG and
intra-CG parallelism for the S-Aligner. Also discussed is our
bit-level encoding strategy and the exploitation of local device
memory.

@@

The highest level of parallelization employs a coarse-grained
partitioning scheme over a block distribution of reads and reference
genome using MPI. In our experiments we process up to 1.6 TB of read
data using up to 13,312 SW26010 nodes executing more than 50,000
dispatched processes. Thus, a na\"{\i}ve partitioning of the read file
into 50,000 pieces of approximately 32 MB size is not a suitable
solution because of the excessive data replication of the reference
genome index. Instead, we employ a partitioning strategy based on a
{\em task grid pattern} for the inter-CG parallelization by
concurrently assigning pairs of reference genome blocks and read
chunks to individual CGs. The two dimensions of the grid are spanned
by the reference block identifiers (rows) and read chunk identifiers
(columns) as shown in Figure \ref{TaskGrid}(a). Thus, processes in the
same row share a unique reference genome (index) block while processes
within a column use the same read chunk. Each process executes several
cells in one row in order to avoid replication of the relatively big
reference index.

@@

The read input file is partitioned into chunks of fixed size. This can
be easily achieved by splitting by lines. For the reference genome, we
first concatenate the individual chromosomes to form one long
string. We then divide it into blocks of suitable size according to
the memory capacity of a CG ($\approx$300 million bps per block). Note
that this partitioning scheme is disadvantageous when a chromosome is
scattered over two blocks or a read is mapped to the junction point of
two chromosomes. This issue can be resolved by padding `N's between
chromosomes. Moreover, when dividing a chromosome into two blocks, we
store several base pairs in a neighborhood around the break point in
both chunks. As a result, a read is always mapped correctly to a
contiguous region. Figure~\ref{RefDiv} depicts this approach.

The computing nodes of Sunway Taihu Light are connected to a network
file system via a 1 GBit/s interface, while the overall file system
bandwidth is $\sim290$ GB/s for the whole cluster. In practice,
efficient data distribution patterns have to be employed to reach
reasonable performance. The file system bandwidth is low compared with
the bisection bandwidth of $\sim70$ TB/s among compute nodes. To
address this issue, we employ a {\em group-and-broadcast} data reuse
strategy to avoid redundant file system accesses by sharing identical
data via Sunway's network among nodes. As tasks are assigned to grid
cells, we cluster processes working on the same reference genome block
to a row group ({\em ref broadcast group}), while processes working on
the same read chunk are clustered to a column group. If a process
calculates cells in the first row or first column ({\em read broadcast
  group}), it loads the corresponding reference block or read chunk
from the file system as shown in
Figure~\ref{TaskGrid}(b). Subsequently, it broadcasts the loaded data
to processes calculating the same row or column, while the other
processes wait for the broadcast data as shown in
Figure~\ref{TaskGrid}(c). After the data-loading stage, processes can
perform their computation concurrently.

\begin{figure}[!htb]
  \includegraphics[width=\linewidth]{figures/AsyncRead}
  \caption{Asynchronous data loading strategy: (a) default case, and
    (b) case when there is no significantly shorter reference block.}
  \label{AsyncRead}
\end{figure}

Our reference genome partitioning strategy illustrated in
Figure~\ref{RefDiv} often generates one reference genome block that is
significantly shorter than the others. This block is assigned to the
first row of our task grid. Thus, processes in the first row take less
time during computation and will have sufficient time to load reads
for the subsequent round of computation. This approach therefore
reduces the idle time of other processes waiting for read data.
Figure \ref{AsyncRead}(a) illustrates this strategy.

We further provide an optional asynchronous data-loading strategy in
case that there is no significantly shorter reference block. In this
case we add a row of processes that loads only reads. When other
processes in the column are computing alignments, these processes
exclusively load new read chunks. After loading and computation have
been completed, all processes within a column receive the loaded data
by means of an \texttt{MPI\textunderscore Bcast}. In our experiments,
this approach reduces the idle time between computing two read chunks
by over 90\% when using 8 processes. Figure \ref{AsyncRead}(b)
illustrates this strategy.

@@

The second level of parallelism exploits the threading capabilities of
a CG. Our design within a single node is a comprehensive read mapper
based on the described seed-and-extend approach using the three-stage
pipeline: filtration, verification, and alignment. The filtration step
uses a look-up table of $q$-grams of the assigned reference genome
block that can be loaded from a preprocessed index file or
alternatively calculated on-the-fly by a radix sort-based hash table
construction method. After the MP performs the lookup of each
non-overlapping $q$-gram of a read, it stores the retrieved intervals
in a SIMD-friendly manner allowing for efficient vectorization during
the subsequent verification step on the SPs. Verification selects all
intervals that comply with the restricted edit distance using Myers'
bit-parallel algorithm. Afterwards, we employ a banded version of the
Smith-Waterman algorithm on the SPs to align the remaining intervals
that have passed verification. Figure~\ref{FrmWk} shows the workflow
of our intra-CG implementation.

@@

The intra-CG parallelization is realized by a task parallelization
scheme implemented using spawn and join calls of the {\em athread}
library. The MP executes the filtration stage while the SPs process
the verification step. Each read is divided into non-overlapping
$q$-grams. The MP looks up those $q$-grams in the reference genome
index and returns a number of intervals. The MP copies them to a
buffer that is transferred to the LDM of SPs for verification. We
employ a dual buffer strategy by allocating two buffers for storing
the intervals identified by the MP. When the MP has finished filling
one buffer, it waits for the SP thread group to join, and then
dispatches an SP thread group to verify these intervals. Subsequently,
the MP fills another buffer, while the SP thread group performs
verification. These steps are repeated until all reads are processed
(see Figure \ref{AsyncPara}). The dual buffer strategy reduces the
idle times of SPs and is also used for implementing the subsequent
alignment stage.

@@

Figure~\ref{SPParallel} illustrates our framework for dividing the
computation between the MP and SPs within a single CG. The number of
seeds identified for a given read can vary while the verification time
of a single seed is constant. Thus, a simple static assignment of seed
intervals per read to SPs can cause workload imbalance.  To balance
the workload, we introduce the parameter $slice\_num$, which restricts
the maximum number of seed intervals to be verified for one read. If
there are more than $slice\_num$ seed intervals to be verified for a
read, we spawn verification threads with the first $slice\_num$
intervals. The remaining intervals will be verified in a subsequent
round of spawns. A value of $slice\_num$ that is too large can reduce
the overhead of spawning SP thread groups, but it worsens workload
balance, and vice versa. Our default value of $slice\_num = 100$
provides a good trade-off in practice.

@@

Implementing Myers' bit-parallel algorithm using ShenWei's SIMD
intrinsics requires a bit-level encoding of DNA sequences to
efficiently evaluate the characteristic function $\chi(s_{i-1} =
s'_{j-1})$ for all $i, j$.

Different approaches have been used in existing aligners: BWA
\cite{bwa} employs an {\em array of structures} (AoS) while RazerS3
\cite{razers3} stores a DNA sequence in a profile using one-hot
encoding. While the AoS approach is more space-efficient and
cache-friendly, the usage of a profile with one-hot encoding supports
bit-parallel computation in a more efficient way.  Because of the
small size of the LDM and the requirement of bit parallelism, we
employ an encoding strategy that combines an AoS with an SoA ({\em
  structure of arrays}) approach, which we call an AoSoA ({\em array
  of structures of arrays}). This mixed strategy builds two
bit-vectors in an interleaved fashion. Thus, we can fetch the encoded
sequence by one DMA-intrinsic and calculate $\chi(s_{i-1} = s'_{j-1})$
in terms of two logical equality operations. Figure~\ref{MixPack}
illustrates the encoding strategy.

\begin{figure}[!htb]
  \includegraphics[width=0.9\linewidth]{figures/MixPack}
  \caption{Bit-level encoding strategies for DNA sequences. For
    simplicity, we use a 4-bit word instead of 64-bit word for
    illustrating our combined AoSoA strategy (bottom).}
  \label{MixPack}
\end{figure}

When updating the five bit-vectors according to
Equation~\ref{one-hot-myers} along the columns in a bit-parallel
fashion, a circular dependency needs to be resolved: $D^0_{i,j}$
depends on $H^-_{i-1, j}$, which in turn depends on $D^0_{i-1,j}$ (a
value we have not computed yet). Myers \cite{myers} has shown that
this can be solved by using logical operations and an
addition. ShenWei's SIMD instructions support 256-bit integer
arithmetic, such as \texttt{simd\_uaddo\_take\_carry} (short for SIMD
unsigned octa-word add, taking carry) which adds two 256-bit operands
and returns a 256-bit unsigned integer. Thus, different from
implementations of Myers' algorithm on other architectures (e.g.,
\cite{chacon}), our implementation does not require additional
instructions to process carries generated within a SIMD
lane. Unfortunately, there is no 256-bit compare instruction. Thus, we
have implemented comparisons in terms of shifting operations.

@@

Here we use a busy-waiting strategy because the time used for DMA
fetching is usually shorter than the time required for
verification. Thus, in the common case the DMA reply word needs to be
checked just once in order to pass the busy-wait loop. Our
experimental results show that our asynchronous data transfer
implementation can almost hide the DMA-fetching latency completely and
gains a performance improvement of a factor of 22 compared with an
implementation based on \texttt{memcpy}.

>>>>

This section describes the implementation of both the inter-CG and
intra-CG parallelism for S-Aligner. Also discussed are our
bit-level encoding strategy and the exploitation of local device
memory.

@@

The highest level of parallelization employs a coarse-grained
partitioning scheme over a block distribution of reads and reference
genomes using MPI. In our experiments we process up to 1.6 TB of read
data using up to 13,312 SW26010 nodes executing more than 50,000
dispatched processes. Thus, a na\"{\i}ve partitioning of the read file
into 50,000 pieces of approximately 32 MB is not a suitable
solution because of the excessive data replication of the reference
genome index. Instead, we employ a partitioning strategy based on a
{\em task grid pattern} for the inter-CG parallelization by
concurrently assigning pairs of reference genome blocks and read
chunks to individual CGs. The two dimensions of the grid are spanned
by the reference block identifiers (rows) and read chunk identifiers
(columns) as shown in Figure \ref{TaskGrid}(a). Thus, processes in the
same row share a unique reference genome (index) block while processes
within a column use the same read chunk. Each process executes several
cells in one row in order to avoid replicating the relatively big
reference index.

@@

The read input file is partitioned into chunks of fixed size. This partitioning can
be easily achieved by splitting by lines. For the reference genome, we
first concatenate the individual chromosomes to form one long
string. We then divide it into blocks of suitable size according to
the memory capacity of a CG ($\approx$300 million bps per block). Note
that this partitioning scheme is disadvantageous when a chromosome is
scattered over two blocks or a read is mapped to the junction point of
two chromosomes. This issue can be resolved by padding N's between
chromosomes. Moreover, when dividing a chromosome into two blocks, we
store several base pairs in a neighborhood around the break point in
both chunks. As a result, a read is always mapped correctly to a
contiguous region. Figure~\ref{RefDiv} depicts this approach.

The computing nodes of Sunway Taihu Light are connected to a network
file system via a 1 GBit/s interface, while the overall file system
bandwidth is $\sim290$ GB/s for the whole cluster. In practice,
efficient data distribution patterns must be employed in order to reach
reasonable performance. The file system bandwidth is low compared with
the bisection bandwidth of $\sim70$ TB/s among compute nodes. To
address this issue, we employ a {\em group-and-broadcast} data reuse
strategy to avoid redundant file system accesses by sharing identical
data via Sunway's network among nodes. As tasks are assigned to grid
cells, we cluster processes working on the same reference genome block
to a row group ({\em ref broadcast group}), while processes working on
the same read chunk are clustered to a column group. If a process
calculates cells in the first row or first column ({\em read broadcast
  group}), it loads the corresponding reference block or read chunk
from the file system as shown in
Figure~\ref{TaskGrid}(b). Subsequently, it broadcasts the loaded data
to processes calculating the same row or column, while the other
processes wait for the broadcast data as shown in
Figure~\ref{TaskGrid}(c). After the data-loading stage, processes can
perform their computation concurrently.

\begin{figure}[!htb]
  \includegraphics[width=\linewidth]{figures/AsyncRead}
  \caption{Asynchronous data loading strategy: (a) default case and
    (b) case when there is no significantly shorter reference block.}
  \label{AsyncRead}
\end{figure}

Our reference genome-partitioning strategy illustrated in
Figure~\ref{RefDiv} often generates one reference genome block that is
significantly shorter than the others. This block is assigned to the
first row of our task grid. Thus, processes in the first row take less
time during computation and will have sufficient time to load reads
for the subsequent round of computation. This approach therefore
reduces the idle time of other processes waiting for read data.
Figure \ref{AsyncRead}(a) illustrates this strategy.

We further provide an optional asynchronous data-loading strategy in
case no significantly shorter reference block exists. In this
case we add a row of processes that loads only reads. When other
processes in the column are computing alignments, these processes
exclusively load new read chunks. After loading and computation have
been completed, all processes within a column receive the loaded data
by means of an \texttt{MPI\textunderscore Bcast}. In our experiments,
this approach reduces the idle time between computing two read chunks
by over 90\% when using 8 processes. Figure \ref{AsyncRead}(b)
illustrates this strategy.

@@

The second level of parallelism exploits the threading capabilities of
a CG. Our design within a single node is a comprehensive read mapper
based on the described seed-and-extend approach using a three-stage
pipeline: filtration, verification, and alignment. The filtration step
uses a lookup table of $q$-grams of the assigned reference genome
block that can be loaded from a preprocessed index file or,
alternatively, calculated on the fly by a radix sort-based hash table
construction method. After the MP performs the lookup of each
non-overlapping $q$-gram of a read, it stores the retrieved intervals
in a SIMD-friendly manner allowing for efficient vectorization during
the subsequent verification step on the SPs. Verification selects all
intervals that comply with the restricted edit distance using Myers'
bit-parallel algorithm. We then employ a banded version of the
Smith-Waterman algorithm on the SPs to align the remaining intervals
that have passed verification. Figure~\ref{FrmWk} shows the workflow
of our intra-CG implementation.

@@

The intra-CG parallelization is realized by a task parallelization
scheme implemented by using spawn and join calls of the {\em athread}
library. The MP executes the filtration stage while the SPs process
the verification step. Each read is divided into non-overlapping
$q$-grams. The MP looks up those $q$-grams in the reference genome
index and returns a number of intervals. The MP copies them to a
buffer that is transferred to the LDM of SPs for verification. We
employ a dual buffer strategy by allocating two buffers for storing
the intervals identified by the MP. When the MP has finished filling
one buffer, it waits for the SP thread group to join and then
dispatches an SP thread group to verify these intervals. Subsequently,
the MP fills another buffer, while the SP thread group performs
verification. These steps are repeated until all reads are processed
(see Figure \ref{AsyncPara}). The dual buffer strategy reduces the
idle times of SPs and is also used for implementing the subsequent
alignment stage.

@@

Figure~\ref{SPParallel} illustrates our framework for dividing the
computation between the MP and SPs within a single CG. The number of
seeds identified for a given read can vary, whereas the verification time
of a single seed is constant. Thus, a simple static assignment of seed
intervals per read to SPs can cause workload imbalance.  To balance
the workload, we introduce the parameter $slice\_num$, which restricts
the maximum number of seed intervals to be verified for one read. If
more than $slice\_num$ seed intervals are to be verified for a
read, we spawn verification threads with the first $slice\_num$
intervals. The remaining intervals will be verified in a subsequent
round of spawns. A value of $slice\_num$ that is too large can reduce
the overhead of spawning SP thread groups, but it worsens the workload
balance, and vice versa. Our default value of $slice\_num = 100$
provides a good trade-off in practice.

@@

Implementing Myers' bit-parallel algorithm by using ShenWei's SIMD
intrinsics requires a bit-level encoding of DNA sequences to
efficiently evaluate the characteristic function $\chi(s_{i-1} =
s'_{j-1})$ for all $i, j$.

Different approaches have been used in existing aligners: BWA
\cite{bwa} employs an {\em array of structures} (AoS), whereas RazerS3
\cite{razers3} stores a DNA sequence in a profile by using one-hot
encoding. While the AoS approach is more space efficient and
cache friendly, the usage of a profile with one-hot encoding supports
bit-parallel computation in a more efficient way.  Because of the
small size of the LDM and the requirement of bit parallelism, we
employ an encoding strategy that combines an AoS with an SoA ({\em
  structure of arrays}) approach, which we call an AoSoA ({\em array
  of structures of arrays}). This mixed strategy builds two
bit-vectors in an interleaved fashion. Thus, we can fetch the encoded
sequence by one DNA-intrinsic and calculate $\chi(s_{i-1} = s'_{j-1})$
in terms of two logical equality operations. Figure~\ref{MixPack}
illustrates the encoding strategy.

\begin{figure}[!htb]
  \includegraphics[width=0.9\linewidth]{figures/MixPack}
  \caption{Bit-level encoding strategies for DNA sequences. For
    simplicity, we use a 4-bit word instead of a 64-bit word for
    illustrating our combined AoSoA strategy (bottom).}
  \label{MixPack}
\end{figure}

When updating the five bit-vectors according to
Equation~\ref{one-hot-myers} along the columns in a bit-parallel
fashion, a circular dependency needs to be resolved: $D^0_{i,j}$
depends on $H^-_{i-1, j}$, which in turn depends on $D^0_{i-1,j}$ (a
value we have not computed yet). Myers \cite{myers} has shown that
this can be solved by using logical operations and an
addition. ShenWei's SIMD instructions support 256-bit integer
arithmetic, such as \texttt{simd\_uaddo\_take\_carry} (short for SIMD
unsigned octa-word add, taking carry), which adds two 256-bit operands
and returns a 256-bit unsigned integer. Thus, different from
implementations of Myers' algorithm on other architectures (e.g.,
\cite{chacon}), our implementation does not require additional
instructions to process carries generated within a SIMD
lane. Unfortunately, there is no 256-bit compare instruction. Thus, we
have implemented comparisons in terms of shifting operations.

@@

Here we use a busy-waiting strategy because the time used for DMA
fetching is usually shorter than the time required for
verification. Thus, in the common case the DMA reply word needs to be
checked just once in order to pass the busy-wait loop. Our
experimental results show that our asynchronous data transfer
implementation can hide the DMA-fetching latency almost completely and
gains a performance improvement of a factor of 22 compared with an
implementation based on \texttt{memcpy}.

### 2017-04-07 text/evaluation.tex
minor edits throughout

In this experiment, we used the first chromosome of GRCh38 as
reference and 20K reads of length 200 bps generated by Mason as input
data.

First, we evaluated the runtime performance of four different
implementations of edit distance calculation using Myers' algorithm in
one CG: (1) a na\"ive method using the MP only, (2) a na\"ive method
with a non-vectorized SP parallelization, (3) a vectorized Myers'
algorithm with the MP only, and (4) our vectorized Myers' algorithm
with SP parallelization.

@@

The result shown in Figure \ref{VarVerCha} demonstrates that making
full use of SPs and bit-parallel SIMD vectorization gains dramatic
speedups (118$\times$ and 63$\times$, respectively). Since the MP does
not support high-precision (256-bit integer) extensions, we
implemented a workaround for the high-precision addition
operation. This makes it run slower on a single MP than on a single
SP. Thus, using SPs can gain a superlinear speedup in this
application. Furthermore, our bit-parallel SIMD implementation is able
to update 256 cells simultaneously using 23 instructions (i.e., 0.09
instructions per cell) while the na\"ive implementation computes one
cell using 6 instructions. This results in a theoretical speedup of
$\sim$67$\times$, which explains the actually achieved speedup of
63$\times$.

Second, we evaluated the impact of using asynchronous filtration and
the impact of using DMA intrinsics versus \texttt{memcpy}. The result
is shown in Figure \ref{VarOptCha}.

@@

The result demonstrates that DMA fetching is key for achieving high
performance when using SPs, since a speedup of $\sim$22$\times$ is
gained over \texttt{memcpy}. This can be attributed to two factors:
(1) DMA transfers require fewer compute resources since they are
performed by the memory controller, whereas \texttt{memcpy} uses an SP
to write data to memory; (2) DMA fetching can be performed
asynchronously, thus enabling the latency of fetching data to be
hidden by computation. Furthermore, asynchronous filtration gains
$\sim$15\% speedup.

@@

In this experiment, we use the first chromosome from GRCh38 as
reference and various number of reads from ERR013135. The runtime of
S-Aligner on a single SW26010 node is compared with RazerS3, a
representative all-mapper. Since RazerS3 does not support ShenWei's
architecture, we ran it on a machine with Intel processors using
multithreading with default parameters (e.g., the accuracy parameter
is set to 98\%). Measured runtimes are shown in Table
\ref{SingleNode}.  One can see that S-Aligner executed on a single
SW26010 (260 cores running at 1.45 GHz) outperforms RazerS3 running on
eight Xeon E7-8860v3 CPUs (128 cores running at 3.2 GHz).

\begin{table}
  \begin{threeparttable}
    \caption{Runtime comparison between S-Aligner running on a single
      CG and RazerS3 running on a single node with eight Xeon
      E7-8860v3 CPUs.}
    \label{SingleNode}
    \begin{tabular}{@{\extracolsep{2pt}}rrrrr}
      \hline
      \multicolumn{1}{c}{Ref} &
      \multicolumn{2}{c}{Reads} &
      \multicolumn{2}{c}{Time(s)}\\
      \cline{2-3}
      \cline{4-5}
      \multicolumn{1}{c}{\#bps} &
      \multicolumn{1}{c}{Count} &
      \multicolumn{1}{c}{\#bps} &
      \multicolumn{1}{c}{RazerS3\tnote{\textdagger}} &
      \multicolumn{1}{c}{S-Aligner\tnote{\textdaggerdbl}}\\
      \hline
      116M & 40M & 108 &  939 & 405\\
      253M & 40M & 108 &  3,044 & 892\\
      253M & 40M & 200 &  3,430 & 2,580\\
      \hline
    \end{tabular}
    \begin{tablenotes}
    \item[\textdagger] result is from a machine with eight Xeon
      E7-8860v3 (128 cores up to 3.20GHz).
    \item[\textdaggerdbl] result is from a node with a SW26010 (260
      cores with frequency of 1.45GHz).
    \end{tablenotes}
  \end{threeparttable}
\end{table}

Rabema \cite{rabema} (Read Alignment BEnchMark) is a well-defined read
alignment benchmark that can evaluate the quality of read mappers in
both {\em all} and {\em any-best} mode. We evaluated the accuracy of
S-Aligner in all mode based on Rabema by using the result of RazerS3
with the accuracy parameter set to 100\% as gold standard. S-Aligner
can find up to 99.8\% of the normalized intervals found by RazerS3
(100\% accuracy), which is higher than the accuracy of RazerS3
executed with default parameters (98\% accuracy). Detailed results are
provided in Table~\ref{AccuEval}.

@@

To evaluate weak and strong scalability, we measured the runtimes of
S-Aligner using different numbers of nodes ranging from 13 to
13,312. The full GRCh38 assembly of the human genome was used as
reference. Input reads were simulated by Mason with an error rate of
4\% and a length of 200 bps. Weak scalability was measured for numbers
of nodes ranging from 13 to 3,328 by increasing the number of input
reads from 16 million to 4 billion, correspondingly. Strong
scalability was measured by increasing the number of nodes from 3,328
to 13,312 while keeping the number of input reads constant at 4
billion. We measured the performance in terms of {\em read base-pairs
  processed per second} (bpps) in total and normalized per node (see
Table \ref{paraexp}). Figure \ref{VarProcCha} shows the achieved
speedups and efficiencies. The results demonstrate weak scalability
since the normalized node performance is almost constant for the
number of nodes ranging from 13 to 3,328. Furthermore, the node
performance only slightly decreases when increasing the number of
nodes from 3,328 to 13,312 while keeping the amount of input reads
constant at 4 billion, thus demonstrating strong scalability for
sufficiently large input datasets.

\begin{table}[!htb]
  \begin{threeparttable}
    \caption{Performance and runtime evaluation of S-Aligner using
      different numbers of nodes.}
    \label{paraexp}
    \begin{tabular}{rrrrrr}
      \hline
      \multicolumn{1}{c}{\multirow{2}{*}{\#nodes}} &
      \multicolumn{2}{c}{Reads}  &
      \multicolumn{1}{c}{\multirow{2}{*}{Time(s)}} &
      \multicolumn{2}{c}{Performance (bpps)} \\
      \cline{2-3}
      \cline{5-6}
      \multicolumn{1}{c}{} &
      \multicolumn{1}{c}{Count} &
      \multicolumn{1}{c}{\#bps} & &
      \multicolumn{1}{c}{Total (M)} &
      \multicolumn{1}{c}{Node (K)}\\
      \hline
      13 & 16M & 200 & 1,315 & 2.43 & 185.43\\
      52 & 64M & 200 & 1,321 & 9.69 & 184.80\\
      208 & 256M & 200 & 1,321 & 38.76 & 184.66\\
      832 & 1,024M & 200 & 1,327 & 154.33 & 184.59\\
      3,328 & 4,096M & 200 & 1,349 & 607.26 & 179.67\\
      13,312 & 4,096M & 200 & 344 & 2,381.40 & 178.90\\
      \hline
    \end{tabular}
    \begin{tablenotes}
    \item
      \#nodes is the number of nodes involved in computing.
    \item
      bpps is short for million bp processed per second.
    \item
      M / K indicates that column is presented in millions / thousands.
    \end{tablenotes}
  \end{threeparttable}
\end{table}

>>>>

In our single-node experiment, we used the first chromosome of GRCh38 as
reference and 20K reads of length 200 bps generated by Mason as input
data.

First, we evaluated the runtime performance of four different
implementations of edit distance calculation using Myers' algorithm in
one CG: (1) a na\"ive method using the MP only, (2) a na\"ive method
with a nonvectorized SP parallelization, (3) a vectorized Myers
algorithm with the MP only, and (4) our vectorized Myers algorithm
with SP parallelization.

@@

The results in Figure \ref{VarVerCha} show that making
full use of SPs and bit-parallel SIMD vectorization gains dramatic
speedups (118$\times$ and 63$\times$, respectively). Since the MP does
not support high-precision (256-bit integer) extensions, we
implemented a workaround for the high-precision addition
operation. This makes it run slower on a single MP than on a single
SP. Thus, using SPs can gain a superlinear speedup in this
application. Furthermore, our bit-parallel SIMD implementation is able
to update 256 cells simultaneously using 23 instructions (i.e., 0.09
instructions per cell) while the na\"ive implementation computes one
cell using 6 instructions. This results in a theoretical speedup of
$\sim$67$\times$, which explains the actually achieved speedup of
63$\times$.

Second, we evaluated the impact of using asynchronous filtration and
the impact of using DMA intrinsics versus \texttt{memcpy}. The results
are shown in Figure \ref{VarOptCha}.

@@

The results demonstrate that DMA fetching is key for achieving high
performance when using SPs, since a speedup of $\sim$22$\times$ is
gained over \texttt{memcpy}. This can be attributed to two factors:
(1) DMA transfers require fewer compute resources since they are
performed by the memory controller, whereas \texttt{memcpy} uses an SP
to write data to memory; and (2) DMA fetching can be performed
asynchronously, thus enabling the latency of fetching data to be
hidden by computation. Furthermore, asynchronous filtration gains
$\sim$15\% speedup.

@@

In our experiment with RazerS3, we used the first chromosome from GRCh38 as
reference and various numbers of reads from ERR013135. The runtime of
S-Aligner on a single SW26010 node was compared with that of RazerS3, a
representative all-mapper. Since RazerS3 does not support ShenWei's
architecture, we ran it on a machine with Intel processors using
multithreading with default parameters (e.g., the accuracy parameter
is set to 98\%). Measured runtimes are shown in Table
\ref{SingleNode}.  One can see that S-Aligner executed on a single
SW26010 (260 cores running at 1.45 GHz) outperforms RazerS3 running on
eight Xeon E7-8860v3 CPUs (128 cores running at 3.2 GHz).

\begin{table}
  \begin{threeparttable}
    \caption{Runtime comparison between S-Aligner running on a single
      CG and RazerS3 running on a single node with eight Xeon
      E7-8860v3 CPUs.}
    \label{SingleNode}
    \begin{tabular}{@{\extracolsep{2pt}}rrrrr}
      \hline
      \multicolumn{1}{c}{Ref} &
      \multicolumn{2}{c}{Reads} &
      \multicolumn{2}{c}{Time(s)}\\
      \cline{2-3}
      \cline{4-5}
      \multicolumn{1}{c}{\#bps} &
      \multicolumn{1}{c}{Count} &
      \multicolumn{1}{c}{\#bps} &
      \multicolumn{1}{c}{RazerS3\tnote{\textdagger}} &
      \multicolumn{1}{c}{S-Aligner\tnote{\textdaggerdbl}}\\
      \hline
      116M & 40M & 108 &  939 & 405\\
      253M & 40M & 108 &  3,044 & 892\\
      253M & 40M & 200 &  3,430 & 2,580\\
      \hline
    \end{tabular}
    \begin{tablenotes}
    \item[\textdagger] Result is from a machine with eight Xeon
      E7-8860v3 (128 cores up to 3.20 GHz).
    \item[\textdaggerdbl] Result is from a node with a SW26010 (260
      cores with frequency of 1.45 GHz).
    \end{tablenotes}
  \end{threeparttable}
\end{table}

Rabema \cite{rabema} (Read Alignment BEnchMark) is a well-defined read
alignment benchmark that can evaluate the quality of read mappers in
both {\em all} and {\em any-best} mode. We evaluated the accuracy of
S-Aligner in {\em all} mode based on Rabema by using the result of RazerS3
with the accuracy parameter set to 100\% as gold standard. S-Aligner
can find up to 99.8\% of the normalized intervals found by RazerS3
(100\% accuracy), which is higher than the accuracy of RazerS3
executed with default parameters (98\% accuracy). Detailed results are
provided in Table~\ref{AccuEval}.

@@

To evaluate weak and strong scalability, we measured the runtimes of
S-Aligner using different numbers of nodes ranging from 13 to
13,312. The full GRCh38 assembly of the human genome was used as
reference. Input reads were simulated by Mason with an error rate of
4\% and a length of 200 bps. Weak scalability was measured for numbers
of nodes ranging from 13 to 3,328 by increasing the number of input
reads from 16 million to 4 billion, correspondingly. Strong
scalability was measured by increasing the number of nodes from 3,328
to 13,312 while keeping the number of input reads constant at 4
billion. We measured the performance in terms of {\em read base-pairs
  processed per second} in total and normalized per node (see
Table \ref{paraexp}). Figure \ref{VarProcCha} shows the achieved
speedups and efficiencies. The results demonstrate weak scalability
since the normalized node performance is almost constant for the
number of nodes ranging from 13 to 3,328. Furthermore, the node
performance only slightly decreases when increasing the number of
nodes from 3,328 to 13,312 while keeping the number of input reads
constant at 4 billion, thus demonstrating strong scalability for
sufficiently large input datasets.

\begin{table}[!htb]
  \begin{threeparttable}
    \caption{Performance and runtime evaluation of S-Aligner using
      different numbers of nodes.}
    \label{paraexp}
    \begin{tabular}{rrrrrr}
      \hline
      \multicolumn{1}{c}{\multirow{2}{*}{\# Nodes}} &
      \multicolumn{2}{c}{Reads}  &
      \multicolumn{1}{c}{\multirow{2}{*}{Time(s)}} &
      \multicolumn{2}{c}{Performance (bpps)} \\
      \cline{2-3}
      \cline{5-6}
      \multicolumn{1}{c}{} &
      \multicolumn{1}{c}{Count} &
      \multicolumn{1}{c}{\# bps} & &
      \multicolumn{1}{c}{Total (M)} &
      \multicolumn{1}{c}{Node (K)}\\
      \hline
      13 & 16M & 200 & 1,315 & 2.43 & 185.43\\
      52 & 64M & 200 & 1,321 & 9.69 & 184.80\\
      208 & 256M & 200 & 1,321 & 38.76 & 184.66\\
      832 & 1,024M & 200 & 1,327 & 154.33 & 184.59\\
      3,328 & 4,096M & 200 & 1,349 & 607.26 & 179.67\\
      13,312 & 4,096M & 200 & 344 & 2,381.40 & 178.90\\
      \hline
    \end{tabular}
    \begin{tablenotes}
    \item
      \# Nodes is the number of nodes involved in computing..
    \item
      bpps is short for million base pairs processed per second.
    \item
      M (K) indicates that column is presented in millions (thousands).
    \end{tablenotes}
  \end{threeparttable}
\end{table}
